sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-05 doc/paper-jss/owls.Rmd
Edits

@@

```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='ht', fig.align='center'}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='ht', fig.align='center'}
qqnorm(pred$resid)
qqline(pred$resid)
```

>>>>

```{r owl-knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  dpi = 125,
  dev = "png",
  optipng = "-strip all"
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

@@

```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='ht', fig.align='center', dev='png', dpi=100}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='ht', fig.align='center', dpi=90, dev='png'}
qqnorm(pred$resid)
qqline(pred$resid)
```

### 2022-11-05 doc/paper-jss/paper-jss.Rmd
Edits

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
output: rticles::jss_article
---

```{r main-setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "png",
  dpi = 180
  # optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "> ")
)
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} or \pkg{inlabru} (Appendix 4) and \pkg{VAST} (Appendix 5).

@@

\begin{figure}
\centering
\includegraphics[width=3.5in]{../../figs/timing-spatial-m1-april11}
\caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions. Lines represent means and ribbons 95\% quantiles across 50 random iterations. Note the log x and y axes. \pkg{VAST} is similar to \pkg{sdmTMB} and so is not shown. \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation. \pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates [@miller2019]. Note that spaMM only fits spatial, not spatiotemporal, models. All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}\label{fig:timing}
\end{figure}

@@

Speed-wise, \pkg{sdmTMB} (and by association \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
Furthermore, although both \pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores.

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
output: rticles::jss_article
---

```{r main-setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "> ")
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

\begin{figure}
\centering
\includegraphics[width=3.5in]{../../figs/timing-spatial-m1-april11}
\caption[]{Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions. Lines represent means and ribbons 95\% quantiles across 50 random iterations. Note the log x and y axes. \pkg{VAST} is similar to \pkg{sdmTMB} and so is not shown. \pkg{inlabru} used the empirical Bayes integration strategy and Gaussian approximation with \texttt{bru\_max\_iter = 1}, and the \texttt{like()} formulation. \pkg{mgcv} used \texttt{bam()}, \texttt{method = `fREML'}, and discretized covariates \citep{miller2019}. Note that spaMM only fits spatial, not spatiotemporal, models. All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.}\label{fig:timing}
\end{figure}

@@

Speed-wise, \pkg{sdmTMB} (and due to the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
Furthermore, although both \pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores.
